---
title: "Bankruptcy Prediction"
author: "The Kumar Kids"
date: "2023-04-04"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: spacelab
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Proposal
For our proposal we will be doing the following:

- Using a bankruptcy data set 
- Asking the question whether or not these banks will go bankrupt
- Preliminary explorations are below
- Link to repository: https://github.com/rtomdavs/TO628

# Data Cleaning

## Loading Data and Removing Columns
```{r}
bankruptcy <- read.csv("bankruptcy.csv")
bankruptcy$Net.worth.Assets <- NULL
bankruptcy$Current.Liability.to.Liability <- NULL
bankruptcy$Current.Liability.to.Equity <- NULL
bankruptcy$Net.Income.Flag <- NULL
```


## Normalizing and Dummy-fying the Data
```{r}
# Using model.matrix to convert all the factors to dummy variables
# We are converting all of the factors into dummy variables as the input into knn has to be numeric
bankmm <- as.data.frame(model.matrix(~.-1,bankruptcy))

# Randomize the rows in the data (shuffling the rows)
set.seed(12345)
bank_random <- bankmm[sample(nrow(bankmm)),]

#Normalize the data
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

# we are going to normalize everything 
bank_norm <- as.data.frame(lapply(bank_random, normalize))

```

## Creating our Train and Test Sets
```{r}
# Selects a random subset of 20% of the rows for test data
set.seed(12345)
test_set <- sample(1:nrow(bank_norm), 0.2*nrow(bank_norm)) 

# Create a train set and test set
#First the predictors - all columns except the bankrupt (our y variable) column
bank_train <- bank_norm[-test_set, -match("Bankrupt.",names(bank_norm))]
bank_test <- bank_norm[test_set, -match("Bankrupt.",names(bank_norm))]

#Now the response (aka Labels) - only the bankrupt column
bank_train_labels <- bank_norm[-test_set, "Bankrupt."]
bank_test_labels <- bank_norm[test_set, "Bankrupt."]

#Creating train and test sets with the bankrupt column for other models
bank_train_full<- bank_norm[-test_set, ]
bank_test_full<- bank_norm[test_set, ]
```

# Individual Models and Testing

## Logistic Regression Model Creation
```{r}
logmodel <- lm(Bankrupt. ~., data = bank_train_full)
```

## Logistic Model Testing
```{r}
library(caret)

LogPred <- predict(logmodel, bank_test_full)

predbin_log <- ifelse(LogPred >= 0.5, 1, 0)
confusionMatrix(as.factor(predbin_log), as.factor(bank_test_full$Bankrupt.), positive = "1")
```

## Simple Neural Net Creation
```{r, cache=TRUE}
library(neuralnet)
simplemodel <- neuralnet(Bankrupt. ~ ., data = bank_train_full, hidden = 1)
plot(simplemodel)
```

## Testing our First Neural Net
```{r}
simplepred <- predict(simplemodel,bank_test_full)
summary(simplepred)
simplebin <- ifelse(simplepred >= 0.5, 1, 0)
confusionMatrix(as.factor(simplebin),as.factor(bank_test_full$Bankrupt.), positive = "1")

```

## KNN Model Creation
```{r, cache=TRUE}
library(class)
bank_model <- knn(bank_train,bank_test, bank_train_labels, k = 50)
confusionMatrix(as.factor(bank_model), as.factor(bank_test_labels), positive = "1")
```