---
title: "Part 1: College Dropout Data Preprocessing"
author: "The Kumar Kids"
date: "2023-04-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Data ETL (Extract, Transform and Load)

## Data Loading and Exploration
```{r}
#Reading in the dataset
drops <- read.csv("dataset.csv")
```

```{r}
# get summary stats for all features
summary(drops)
```

## Data Cleaning
### Factorize Categorical Features
```{r}
#Converting appropriate columns into factor
factorcols <- c("Marital.status", "Application.mode", "Daytime.evening.attendance", "Course", "Previous.qualification",
         "Nacionality", "Mother.s.qualification", "Father.s.qualification", "Mother.s.occupation",
         "Father.s.occupation", "Displaced", "Educational.special.needs", "Debtor", "Tuition.fees.up.to.date",
         "Gender", "Scholarship.holder", "International")

drops[factorcols] <- lapply(drops[factorcols], factor)
str(drops)
```


### Split Training Data Set
- "Enrolled" students are still in school and we do not know whether they will drop out
- Therefore, we will extract "Enrolled" students as a separate dataset, train models on historical student data, and apply the best performing model on "Enrolled" students
```{r}
# Removing the "Enrolled" students for later prediction
nonenr <- subset(drops, drops$Target != "Enrolled")

# Saving "Enrolled" students as the prediction set
enr <- subset(drops, drops$Target == "Enrolled")

# Factorizing target
nonenr$Target <- as.factor(nonenr$Target)

# Remove Target variable for prediction dataset
enr$Target <- NULL

str(nonenr)
```


### Create Dummy Variables for Categorical Features
```{r}
#Converting to Model Matrix
dummydrop_ <- as.data.frame(model.matrix(~ . - 1, nonenr))

summary(dummydrop_)
```


#### Issue 1
For categorical features with more than 2 levels, some levels might have few rows of data, while others have a lot. If we convert levels with few data to dummies, the dummies would only add noise when training model.

**Our solution**
For each categorical feature with this problem, we will group the trivial levels to create an aggregated dummy variable called "Others". We define any levels with less than 5% of total values, or any levels other than the top 3 levels with most values (if these is only one level with more than 5%).

```{r, warning=FALSE}
library(dplyr)

# for outputs below, "n" stands for count of the corresponding level, "percentage" indicates the percentage that this level takes out of number of rows

# Marital Status
nonenr %>% count(Marital.status) %>% mutate(percentage = n/sum(n))

# Application Mode
nonenr %>% count(Application.mode) %>% mutate(percentage = n/sum(n))

# Course
nonenr %>% count(Course) %>% mutate(percentage = n/sum(n))

# Previous Qualification
nonenr %>% count(Previous.qualification) %>% mutate(percentage = n/sum(n))

# Nacionality
nonenr %>% count(Nacionality) %>% mutate(percentage = n/sum(n))

# Mother's Qualification
nonenr %>% count(Mother.s.qualification) %>% mutate(percentage = n/sum(n))

# Father's Qualification
nonenr %>% count(Father.s.qualification) %>% mutate(percentage = n/sum(n))

# Mother's Occupation
nonenr %>% count(Mother.s.occupation) %>% mutate(percentage = n/sum(n))

# Father's Occupation
nonenr %>% count(Father.s.occupation) %>% mutate(percentage = n/sum(n))
```

| Variable Name          | Levels to Convert             |
|------------------------|-------------------------------|
| Marital Status         | 3-6                           |
| Application Mode       | 2-7, 9-11, 13, 15-18          |
| Course                 | Balanced                      |
| Previous Qualification | 2-11, 13, 15-17               |
| Nacionality            | 2-8, 10-13, 15-21             |
| Mother's Qualification | 2, 4-12, 14-21, 24-29         |
| Father's Qualification | 2, 4-13, 15-23, 25, 26, 29-34 |
| Mother's Occupation    | 1, 2, 7, 9, 11-32             |
| Father's Occupation    | 1-3, 12-46                    |

#### Convert Levels to Others 
```{r}
# Marital Status
nonenr <- nonenr %>% mutate(Marital.status = ifelse(Marital.status %in% c(seq(3, 6)), "Others", Marital.status))

# Application Mode
nonenr <- nonenr %>% mutate(Application.mode = ifelse(Application.mode %in% c(seq(2,7), seq(9,11), 13, seq(15,18)), "Others", Application.mode))

# Previous Qualification
nonenr <- nonenr %>% mutate(Previous.qualification = ifelse(Previous.qualification %in% c(seq(2,11), 13, seq(15,17)), "Others", Previous.qualification))

# Nacionality
nonenr <- nonenr %>% mutate(Nacionality = ifelse(Nacionality %in% c(seq(2,8), seq(10,13), seq(15,21)), "Others", Nacionality))

# Mother's Qualification
nonenr <- nonenr %>% mutate(Mother.s.qualification = ifelse(Mother.s.qualification %in% c(2, seq(4,12), seq(14,21), seq(24,29)), "Others", Mother.s.qualification))

# Father's Qualification
nonenr <- nonenr %>% mutate(Father.s.qualification = ifelse(Father.s.qualification %in% c(2, seq(4,13), seq(15,23), 25, 26, seq(29,34)), "Others", Father.s.qualification))

# Mother's Occupation
nonenr <- nonenr %>% mutate(Mother.s.occupation = ifelse(Mother.s.occupation %in% c(1, 2, 7, 9, seq(11,32)), "Others", Mother.s.occupation))

# Father's Occupation
nonenr <- nonenr %>% mutate(Father.s.occupation = ifelse(Father.s.occupation %in% c(seq(1,3), seq(12,46)), "Others", Father.s.occupation))
```


```{r}
# Converting transformed data set to Model Matrix again
dummydrop <- as.data.frame(model.matrix(~ . - 1, nonenr))

summary(dummydrop)
```


#### Issue 2
We have observed that `model.matrix()` creates a dummy for all N levels of the categorical feature `Marital Status`, while we only need N-1 levels of dummies to avoid the problem of multicollinearity. 

**Our solution**
Remove a level for these categorical features manually 
```{r}
# marital status has 6 levels, and we have 6 dummies now
dummydrop$Marital.status1 <- NULL

str(dummydrop)
```


### Create Normalized Version of Data
```{r}
# Randomize the rows in the data (shuffling the rows)
set.seed(12345)
randomdrop <- dummydrop[sample(nrow(dummydrop)),]

# Function to normalize a column
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

# Normalize everything 
drop_norm <- as.data.frame(lapply(randomdrop, normalize))

# # Check normalized data
# summary(drop_norm)
```

### Clean Prediction Dataset
Feature engineering: the same transformations are applied on the prediction dataset `enr`
```{r}
# Marital Status
enr <- enr %>% mutate(Marital.status = ifelse(Marital.status %in% c(seq(3, 6)), "Others", Marital.status))

# Application Mode
enr <- enr %>% mutate(Application.mode = ifelse(Application.mode %in% c(seq(2,7), seq(9,11), 13, seq(15,18)), "Others", Application.mode))

# Previous Qualification
enr <- enr %>% mutate(Previous.qualification = ifelse(Previous.qualification %in% c(seq(2,11), 13, seq(15,17)), "Others", Previous.qualification))

# Nacionality
enr <- enr %>% mutate(Nacionality = ifelse(Nacionality %in% c(seq(2,8), seq(10,13), seq(15,21)), "Others", Nacionality))

# Mother's Qualification
enr <- enr %>% mutate(Mother.s.qualification = ifelse(Mother.s.qualification %in% c(2, seq(4,12), seq(14,21), seq(24,29)), "Others", Mother.s.qualification))

# Father's Qualification
enr <- enr %>% mutate(Father.s.qualification = ifelse(Father.s.qualification %in% c(2, seq(4,13), seq(15,23), 25, 26, seq(29,34)), "Others", Father.s.qualification))

# Mother's Occupation
enr <- enr %>% mutate(Mother.s.occupation = ifelse(Mother.s.occupation %in% c(1, 2, 7, 9, seq(11,32)), "Others", Mother.s.occupation))

# Father's Occupation
enr <- enr %>% mutate(Father.s.occupation = ifelse(Father.s.occupation %in% c(seq(1,3), seq(12,46)), "Others", Father.s.occupation))
```

Dummify categorical features
```{r}
# Create dummies
dummy_enr <- as.data.frame(model.matrix(~ . -1, enr))

# drop one level in Marital Status to reduce multicollinearity
dummy_enr$Marital.status1 <- NULL
str(dummy_enr)
```

Normalize all numerical features
```{r}
# Randomize the rows in the data (shuffling the rows)
set.seed(12345)
random_enr <- dummy_enr[sample(nrow(dummy_enr)),]

# Normalize everything 
enroll_norm <- as.data.frame(lapply(random_enr, normalize))

# Check normalized data to make sure every feature is between 0 and 1
summary(enroll_norm)
```


## Save Cleaned Datasets

```{r}
# training set
write.csv(drop_norm, file = "cleaned_nonenrolls.csv", row.names = FALSE)
# prediction set 
write.csv(enroll_norm, file = "cleaned_enrolls.csv", row.names = FALSE)
```

- `cleaned_nonenrolls.csv` will be used for modeling
- `cleaned_enrolls.csv` will be used to apply the prediction models
